---
title: "Adult Mortality in the Metropolis of London 1100–1850"
subtitle: "Supplement: Code structure, data source and processing"
author: 
 - Nils Müller-Scheeßel^[Institute for Pre- and Proto History - Kiel University nils.mueller-scheessel@ufg.uni-kiel.de]
 - Katharina Fuchs^[Institute of Clinical Molecular Biology - Kiel University, k.fuchs@ikmb.uni-kiel.de]
 - Christoph Rinne^[Institute for Pre- and Proto History - Kiel University, crinne@ufg.uni-kiel.de]
date: "`r format(Sys.time(), '%d. %B %Y')`"
license: "CC-BY 4.0"
header-includes:
  #\renewcommand{\contentsname}{Inhalt}
  #\renewcommand{\figurename}{Abb.}
  #\renewcommand{\tablename}{Tab.}
papersize: a4
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
#  pdf_document:
#    number_sections: true
#    toc: true
#    toc_depth: 2
#    df_print: kable
#    extra_dependencies: ["flafter"]
#  html_notebook:
#        code_folding: hide
knit rmdfig_caption: true
urlcolor: blue
link-citations: true # make citations hyperlinks
linkcolor: blue
bibliography: References.bib
csl: offa.csl
---

```{r knitr global options, include=FALSE}
# Global options for knitr
knitr::opts_chunk$set(echo=TRUE, fig.align="center")
```

# Prerequisites {-}

The calculations were made in R using R-Studio. The structure of the code is essentially based on the structure of the text. The raw code is in the file ```order_of_code.R```. The file extended with Markdown is ```order_of_code-doc.RMD``` and the file ```order_of_code-doc.html``` is generated from it.

Note: The base path for rmd files is the folder in which they are located, not the r-project. Consequently, ```order_of_code.R``` and ```order_of_code-doc.RMD``` are both located in the root folder of the project.

> Depending on the hardware, the subsequent code can run for several hours or even a few days.

> Install "Just Another Gibbs Sampler" (JAGS) if you want to run the Bayesian analyses anew.
> Version 4.3 - as used here - can be downloaded in pre-compiled form for a number of OS here: https://sourceforge.net/projects/mcmc-jags/ . 
> The manual can be found here: https://people.stat.sc.edu/hansont/stat740/jags_user_manual.pdf

The code makes extensive use of the function ```source``` to call external code. Thus, the main part of the code remains slim, well structured and readable. 

Install required packages, set some options and link the sources for the helper functions. 

Remark: Depending on your R version the package osmplotr may be installed from github using ```devtools::install_github ("ropensci/osmplotr")```. 

```{r Packages and sources, echo=TRUE}
require(pacman) || install.packages("pacman")
pacman::p_load(coda, cowplot, demogR, dplyr, flexsurv, ggplot2, ggrepel, 
               gridExtra, HMDHFDplus, kableExtra, Metrics, mortAAR, osmdata, 
               psych, readxl, reshape2, rjags, runjags, sf, tidyr, ggspatial)

options(scipen = 999)
options(dplyr.summarise.inform = FALSE)

source("./functions/bayes_cat_poisson.R")
source("./functions/gomp_MLE.R")
source("./functions/gomp_MLE_adapted.R")
source("./functions/gomp_MLE_interval.R")
source("./functions/gomp_anthr_age.R")
source("./functions/gomp_anthr_age_r.R")
source("./functions/gomp_known_age.R")
source("./functions/gomp_known_age_r.R")
source("./functions/helper_functions.R")
source("./functions/lt_MC.R")
source("./functions/lt_MC_Gomp.R")
RNGkind("L'Ecuyer-CMRG") # conservative random number generator to avoid periodicity
```
Important for saving time:  Decide to run extensive code anew (app. 6 h +). In addition, you can set the folder for preprocessed files.

```{r rund code anew, echo=TRUE}
runCodeNew <- FALSE
#runCodeNew <- TRUE

# Ask for credentials of the Human Mortality Database if the code runs anew
if (runCodeNew){
  HMD_username <- readline(prompt = "Enter username: ")
  tstHMD_password <- readline(prompt="Enter password: ")
  credentials <- c(HMD_username, HMD_password)
}

# Specify filename prefix for saved files and create a folder if needed:
saveFileDir = "preprocessed_files"
if (saveFileDir %in% list.files(getwd())) {
  # Dir exists
}else{
  dir.create(file.path(".", saveFileDir), showWarnings = FALSE )
}
```

\newpage
# Chapter 01 Introduction

Figure 1: Exemplary life table curves generated by Gompertz functions with different $\beta$ parameters.

```{r Fig. 1: Gompertz distribution, echo=TRUE, warning=FALSE}
source("./chapter_01_introduction/gompertz_distribution.R")
```

\newpage
# Chapter 02 Materials and methods

Figure 3: Hazard curve (mx) for HMD UK data of the year 1841.

```{r Fig. 3: Hazard curve, echo=TRUE, warning=FALSE}
source("./chapter_02_materials_and_methods/hazard_curve.R")
```

\newpage
# Chapter 03 Data

Figure 4: Major cemeteries in Greater London 1100–1850 used in the present study. 

```{r Fig. 4: Map of London, echo=TRUE, warning=FALSE}
source("./chapter_03_data/London_places.R") 
```

Table 1: Overview of included cemeteries with osteoarchaeological data.

```{r Table 1: Overview of included cemeteries, echo=TRUE, warning=FALSE, results="asis"}
read.table("chapter_03_data/london_cemeteries.txt", header = T, sep = "\t") %>%
  knitr::kable(., caption = "Overview of included cemeteries with osteoarchaeological data.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") %>% unclass() %>% cat()
```

\newpage
Figure 5: Population development of London, compiled from @ref_268519, 39 table 1; @ref_90775, 41; 179 table 5.7; @ref_115940, 655--657.

```{r Fif. 5: Population development, echo=TRUE, warning=FALSE}
source("./chapter_03_data/London_population.R")
grid::grid.newpage()
grid::grid.draw(rbind(london_pop1, london_pop2))
```

Footnote 6: Re-calculation of population increase rates of London from @ref_278969.
Calculated in ./chapter_03_data/London_population.R

```{r Footnote 6: Razzell/Spence 2007, warning=FALSE}
knitr::kable(razz_df, caption = "Re-calculation of population increase rates of London from Razzell/Spence 2007.")%>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

\newpage
# Chapter 04 Results

Preprocessing of data used in figure 6: Estimated modal ages.

## Historical life tables

### Written sources

**Basic statistics and Bayes (extensive) calculations**

The data is referenced and aggregated in "./chapter_04_results/historical_lifetables.R". In this file, all records from individual preprocessing files located in "./liftables_preprocessed/" are ```sourced```.  The corresponding data files are stored in "./data/".

English_Peers.R, russell.txt, Source: @ref_45151, table 2

```{r Basic statistics English Peers, warning=FALSE}
source("./chapter_04_results/historical_lifetables.R")
kable(peers_ranges, caption = "English Peers") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

Medieval_England.R, Christ_church_monks.txt, Source: @ref_54763, 28 table 2

```{r Basic statistics Christ Church monks, warning=FALSE}
kable(monks_ranges, caption = "Christ Church monks") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

London_1728_1840.R, Mortality_bills_1728_1840.txt, Source: @ref_301236, 304 Table 6.5; > 100 years and < 1 year collapsed

```{r Basic statistics Mortality bills 1728-1840 age modes, warning=FALSE}
kable(london_1728_1840_ranges, 
      caption = "London Mortality bills 1728-1840.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

```{r Basic statistics Mortality bills 1728-1840 rate, warning=FALSE}
kable(london_1728_1840_ranges_r, 
      caption = "London Mortality bills 1728-1840, corrected for population growth.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

London_1841_raw_all.R, London_1841_raw.txt, Source: @ref_94043, 19 table q.

```{r Basic statistics census data for London from 1841, warning=FALSE}
kable(London_1841_ranges, 
      caption = "Census data for London from 1841.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

English_Mortality.R, wrigley_et_al_1997_england_1640-1809.txt, Source: @ref_73169, 290 table 6.19

```{r Basic statistics English mortality, warning=FALSE}
kable(eng_mort_ranges, 
      caption = "English mortality data.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

HMD_UK_ranges.R

The data from the Human Mortality Database (https://mortality.org/) were retrieved with a personal account using the R package HMDHFDplus. Therefore, we only provide the processed data here.  

```{r Basic statistics HMD UK, warning=FALSE}
kable(HMD_UK_ranges, caption = "Human Mortality Database UK.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

**Extended statistics**

```{r Etended statistics written sources, warning=FALSE}
kable(peers_result, caption = "English Peers.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(monks_result, caption = "Christ Church monks.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(london_1728_1840_result, 
      caption = "London Mortality bills 1728-1840.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(london_1728_1840_result_r, 
      caption = "London Mortality bills 1728-1840, corrected for population growth.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(London_1841_result, 
      caption = "Census data for London from 1841.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(eng_mort_result, caption = "English mortality data.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(HMD_UK_result, caption = "Human Mortality Database UK.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
```

## London cemeteries

The data is mainly hard coded in the file ./chapter_04_results/Wellcome_DB.R. 

Only St. Bride's crypt is excluded but available from the Museum of London upon request. For general information: [https://www.museumoflondon.org.uk](https://www.museumoflondon.org.uk) go for: Collections > Archaeology at the Museum of London > Wellcome Osteological Research Database > St. Bride's Church Fleet Street. 
If runCodeNew == TRUE  the file ./lifetables_processing/stbrides_crypt.R will ask for the location of the retrieved dataset (Excel sheet) and process the data. In any other case pre-processed data will be loaded. 

```{r cemeteries data pre-processed, warning=FALSE}
source("./lifetables_processing/stbrides_crypt.R")
source("./chapter_04_results/Wellcome_DB.R")
```


```{r wellcome results, warning=FALSE}
kable(wellcome_result, caption = "London cemeteries data.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
kable(wellcome_result_r, caption = "London cemeteries data, corrected for population growth.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
```


## Modal ages from historical and osteological data

Figure 6: Modal ages from historical and osteological data

```{r figure 6 modal ages plot, fig.height=8.5, fig.width=11, echo=TRUE, warning=FALSE}
source("./chapter_04_results/english_wellcome.R")
plot(modal_ages_plot)
```

Table 2: Overview of modelled osteological data from London cemeteries

The data overview is build during pre-processing in ./chapter_04_results/Wellcome_DB.R and saved to a textfile (sep = \\t).

wellcome_overview_all
```{r wellcome_overview_all, warning=FALSE}
kable(wellcome_overview_all, caption = "London cemeteries overview.") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
```


Marylebone.R

Data are hard coded in the code. Sources: @ref_221997, 97--103 table 32 (St Marylebone); @ref_242052, 81 (St Marylebone north of Paddington street)

```{r Basic statistics Marylebone, warning=FALSE}
source("./lifetables_processing/Marylebone.R")
kable(Marylebone_ranges, caption = "St Marylebone, corrected with population growth rate of 2.75 per-cent.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```


Figure 7: St. Bride's Crypt. Density of actual ages and Bayesian model of Gompertz distribution of actual ages and osteological estimates (without correction for population growth).

The plot is build in ./lifetables_processing/stbrides_crypt.R within the if-statement on ```runCodeNew``` (s. data limitations above).

```{r figure 7 St. Bride Crypt, fig.height=4, fig.width=8, warning=FALSE}
plot(stbrides_crypt_plot)
```

Figure 8: Simulation of population increase with known age-at-death and Maximum Likelihood Estimation (MLE) (top four) and osteological estimates, Bayesian model and including rate of increase (bottom four).

```{r Simulation of population increase, warning=FALSE}
source("./chapter_04_results/simulations_pop_incr_run.R")
```

\newpage
# Supplements

The chapter 'Supporting informations' provides details about the London cemeteries included in the study. 

## The Coale & Demeny life tables

Calculation of the lowest $\beta$-value for any of the Coale & Demeny life tables (@ref_22426) which is 0.0391 (the female table "West", level 1).

```{r Coale and Demeny life tables beta-value, warning=FALSE}
source("./chapter_supplement/coale_demeny_life_tables_gompertz.R")
min(gompertz_df$Gompertz_shape)
```

## Simulations

Simulations for evaluation of algorithms for retrieving Gompertz parameters. 
The file "./chapter_supplement/simulations_run.R" provides various tests and plots from the evaluation process. 
The validity of the algorithms is tested via root mean square error (RMSE) of between expected and estimated values for Gompertz $\beta$.

### Known age-at-death

Figure S1: Comparing expected and estimated Gompertz $\beta$-values by different algorithms with known age-at-death (n = 1,000).

```{r Simulations retrieving Gompertz parameters, fig.height=8.5, fig.width=6, warning=FALSE}
source("./chapter_supplement/simulations_run.R")
gridExtra::grid.arrange(grobs = plot_list_shapes, ncol = 3)
```

\newpage

Figure S2: Difference between expected and estimated Gompertz $\beta$-values by different algorithms with known age-at-death (n = 1,000).

```{r Simulations with difference between expected and estimated Gompertz parameters, fig.height=8.5, fig.width=6, warning=FALSE}
source("./chapter_supplement/simulations_run.R")
gridExtra::grid.arrange(grobs = plot_list_diff, ncol = 3)
```

\newpage
Table S1: Root mean square errors (RMSE) for different formulas for fitting known age-at-death, ordered ascendingly.

```{r Simulation evaluation via RMSE}
# table of RMSEs
kable(rmse_result[order(rmse_result$RMSE) ,], caption = "Evaluation of algorithms via RMSE.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") 
```

\newpage
Figure S3: Comparing expected and estimated Gompertz $\beta$-values by different algorithms with estimated age-at-death (n = 1,000).

```{r Simulations retrieving Gompertz parameters with age estimations, fig.height=8.5, fig.width=6, warning=FALSE}
source("./chapter_supplement/simulations_run.R")
gridExtra::grid.arrange(grobs = plot_list_estim_shapes, ncol = 2)
```

\newpage
Table S2: Root mean square errors (RMSE) for different formulas for fitting estimated age-at-death, ordered ascendingly.

```{r Simulation evaluation via RMSE, anthropological categories}
# table of RMSEs
kable(rmse_estim_result[order(rmse_estim_result$RMSE) ,], caption = "Evaluation of algorithms via RMSE for estimated age-at-death.") %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") 
```

Figure S4: Detailed plot for the Bayesian model of expected Gompertz $\beta$-values (left) and sample size (right) vs. difference of expected/estimated Gompertz $\beta$-values.

```{r plot Bayesian model, fig.height=2.5, fig.width=5,warning=FALSE}
#source("./chapter_supplement/simulations_run.R")
gridExtra::grid.arrange(grobs = plot_list_bayes_diff, ncol = 2)
```

\newpage

### Stability of means
Show that means are stable in Bayesian modelling, regardless of the number of MCMC-steps.

```{r Bayesian modeling stable means}
source("./chapter_supplement/bayes_complete.R") # can take a few minutes
kable (bayes_complete, caption = paste0("Bayesian model with simulated dataset and different number of steps. n = ", n, ", Gompertz beta = ", round(beta, 4), ".", sep = "")) %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position","scale_down"))
```

\newpage
# References
